Calibration of P-values for calibration and for deviation of a subpopulation from the full population

### 2022-08-24 tex/paper.tex
also
LaTeX

\documentclass[12pt]{article}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{graphicx}
%\usepackage{enumerate}
\usepackage[authoryear,round]{natbib}
\usepackage{url} % not crucial - just used below for the URL 
\usepackage{setspace}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

@@

@@

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Calibration of P-values for calibration and for deviation
         of a subpopulation from the full population}
  \author{Mark Tygert\\
          Fundamental Artificial Intelligence Research, Meta Platforms, Inc.}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The author's recent research papers,
``Cumulative deviation of a subpopulation from the full population'' and
``A graphical method of cumulative differences between two subpopulations''
(both published in volume 8 of Springer's open-access {\it Journal of Big Data}
during 2021), propose graphical methods and summary statistics,
without extensively calibrating formal significance tests.
The summary metrics and methods can measure the calibration
of probabilistic predictions and can assess differences
in responses between a subpopulation and the full population while controlling
for a covariate or score via conditioning on it.
These recently published papers construct significance tests based
on the scalar summary statistics, but only sketch how to calibrate
the attained significance levels (also known as ``P-values'') for the tests.
The present article reviews and synthesizes work spanning many decades
in order to detail how to calibrate the P-values.
The present paper presents computationally efficient, easily implemented
numerical methods for evaluating properly calibrated P-values,
together with rigorous mathematical proofs guaranteeing their accuracy,
and illustrates and validates the methods with open-source software
and numerical examples.
\end{abstract}

\noindent%
{\it Keywords:} Brownian motion, significance, test, hypothesis,
                numerical methods, graphical methods
\vfill

\newpage
\setstretch{1.5}
%%%%\spacingset{1.5} % DON'T change the spacing!

@@

\begin{theorem}
\label{kucal}
Assume the null hypothesis that the subpopulation has no expected deviation
from the full population (that is, $\E[R_k] = r(S_k)$
for $k = 1$,~$2$, \dots, $n$) and that the third moment obeys
$\E[|R_k - r(S_k)|^3] \le C (v(S_k))^{3/2}$ for $k = 1$,~$2$, \dots, $n$,
where $C$ is a finite positive real number that does not depend on $n$.
Suppose also that the scores $S_1$, $S_2$, \dots, $S_n$ are distinct
for each $n$ and
$\max_{1 \le k \le n} v(S_k) \cdot (W_k)^2 / \sum_{j=1}^n v(S_j) \cdot (W_j)^2$
converges to 0 in the limit as $n$ becomes large.
Then, with $H$ defined in~(\ref{kuiper})
and $\sigma$ defined in~(\ref{general}),
the normalized Kuiper statistic $H/\sigma$
for measuring deviation of a subpopulation from the full population
converges in distribution to the range of the standard Brownian motion
over the unit interval $[0, 1]$.
(The range is the maximum minus the minimum.)
The normalized Kuiper statistic $H/\sigma$ for measuring calibration
converges in distribution to the range of the standard Brownian motion
over the unit interval $[0, 1]$, too, when taking the expected response
at each score to be equal to the score, that is, $r(s) = s$
for every score $s$.
\end{theorem}

@@

\begin{figure}
\begin{center}
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/kuiper.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/kuiper_log.pdf}}
\end{center}
\vspace{-.125in}
\caption{Both plots graph $1 - F(x)$ versus $x$, where
$F$ is defined in~(\ref{cdf}) and is central to Corollary~\ref{corollary}.
The plot on the right uses a logarithmic scale for the vertical axis,
unlike the plot on the left. The vertical dotted line indicates
the value of $x$ corresponding to the mean of the distribution
for which $F$ is the cumulative distribution function.}
\vspace{.5in}
\label{kuiper_plot}
\end{figure}

@@

\begin{figure}
\begin{center}
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/gauss.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/gauss_log.pdf}}
\end{center}
\vspace{-.125in}
\caption{Both plots graph $1 - \Phi(x)$ versus $x$,
where $\Phi$ is the cumulative distribution function
for the standard normal distribution;
$\Phi(x) = \int_{-\infty}^x \exp(-y^2/2) \, dy \, / \, \sqrt{2\pi}$.
The plot on the right uses a logarithmic scale for the vertical axis,
unlike the plot on the left.}
\vspace{-.0625in}
\label{gaussian_plot}
\end{figure}

@@

\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kuiper_ecdf_square.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kuiper_ecdf_sqrt.pdf}}
\end{center}
\vspace{-.125in}
\caption{Calibration curves (empirical cumulative distribution functions
under the null hypothesis of perfectly calibrated data) of the Kuiper P-value
for calibration for sample sizes $n =$ 100, 1,000, 10,000; in each plot,
the dashed line connects the origin $(0, 0)$ to the point $(1, 1)$
and illustrates perfect calibration, while the curve for $n =$ 10,000
is closest to perfect, $n =$ 1,000 is next closest, and $n =$ 100
is the farthest. Subfigure (a) uses scores equispaced
on the unit interval $[0, 1]$,
(b) squares each of the initially equispaced scores,
and (c) takes the square root of each of the initially equispaced scores.
The score $s$ is the predicted probability,
with the expected response $r(s) = s$ to assess calibration.
Each empirical cumulative distribution function plotted arises
from 100,000 data sets generated independently
while assuming the null hypothesis.}
\vspace{-.0625in}
\label{kuc}
\end{figure}

@@

\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kolmogorov_smirnov_ecdf_square.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kolmogorov_smirnov_ecdf_sqrt.pdf}}
\end{center}
\vspace{-.125in}
\caption{Calibration curves (empirical cumulative distribution functions
under the null hypothesis of perfectly calibrated data)
of the Kolmogorov-Smirnov P-value for calibration
for sample sizes $n =$ 100, 1,000, 10,000; in each plot, the dashed line
connects the origin $(0, 0)$ to the point $(1, 1)$ and illustrates
perfect calibration, while the curve for $n =$ 10,000 is closest to perfect,
$n =$ 1,000 is next closest, and $n =$ 100 is the farthest.
Subfigure (a) uses scores equispaced on the unit interval $[0, 1]$,
(b) squares each of the initially equispaced scores,
and (c) takes the square root of each of the initially equispaced scores.
The score $s$ is the predicted probability,
with the expected response $r(s) = s$ to assess calibration.
Each empirical cumulative distribution function plotted arises
from 100,000 data sets generated independently
while assuming the null hypothesis.}
\vspace{-.0625in}
\label{ksc}
\end{figure}

@@

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Python and LaTeX sources:] The Python-package CDEETS contains code
implementing and testing the methods described in the present paper.
The package includes unit tests together with scripts that automatically
reproduce all results and figures reported above. The package also contains
all data sets used above as examples, as well as the LaTeX and BibTeX sources
required for reproducing the paper.

\end{description}

@@

\bibliography{paper}
\end{document}

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{graphicx}
\usepackage[authoryear,round]{natbib}
\usepackage[breaklinks,hidelinks]{hyperref}
\usepackage[margin=1in]{geometry}

@@

\title{Calibration of P-values for calibration and for deviation
       of a subpopulation from the full population}
\author{Mark Tygert\\
        Fundamental Artificial Intelligence Research, Meta Platforms, Inc.\\
        786 Coleman Ave.\ Apt.\ L, Menlo Park, CA 94025-2440\\
        \url{http://tygert.com}\\
        {\tt mark@tygert.com}}

@@

\maketitle

\begin{abstract}
The author's recent research papers,
``Cumulative deviation of a subpopulation from the full population'' and
``A graphical method of cumulative differences between two subpopulations''
(both published in volume 8 of Springer's open-access {\it Journal of Big Data}
during 2021), propose graphical methods and summary statistics,
without extensively calibrating formal significance tests.
The summary metrics and methods can measure the calibration
of probabilistic predictions and can assess differences
in responses between a subpopulation and the full population while controlling
for a covariate or score via conditioning on it.
These recently published papers construct significance tests based
on the scalar summary statistics, but only sketch how to calibrate
the attained significance levels (also known as ``P-values'') for the tests.
The present article reviews and synthesizes work spanning many decades
in order to detail how to calibrate the P-values.
The present paper presents computationally efficient, easily implemented
numerical methods for evaluating properly calibrated P-values,
together with rigorous mathematical proofs guaranteeing their accuracy,
and illustrates and validates the methods with open-source software
and numerical examples.
\end{abstract}

\noindent {\it Keywords:} Brownian motion, significance, test, hypothesis,
                          numerical methods, graph

@@

\begin{theorem}
\label{kucal}
Assume the null hypothesis that the subpopulation has no expected deviation
from the full population (that is, $\E[R_k] = r(S_k)$
for $k = 1$,~$2$, \dots, $n$) and that the third moment obeys
$\E[|R_k - r(S_k)|^3] \le C (v(S_k))^{3/2}$ for $k = 1$,~$2$, \dots, $n$,
where $C$ is a finite positive real number that does not depend on $n$.
Suppose that the scores $S_1$, $S_2$, \dots, $S_n$ are distinct for each $n$
and
$\max_{1 \le k \le n} v(S_k) \cdot (W_k)^2 / \sum_{j=1}^n v(S_j) \cdot (W_j)^2$
converges to 0 in the limit as $n$ becomes large.
Then, with $H$ defined in~(\ref{kuiper})
and $\sigma$ defined in~(\ref{general}),
the normalized Kuiper statistic $H/\sigma$
for measuring deviation of a subpopulation from the full population
converges in distribution to the range of the standard Brownian motion
over the unit interval $[0, 1]$.
(The range is the maximum minus the minimum.)
The normalized Kuiper statistic $H/\sigma$ for measuring calibration
converges in distribution to the range of the standard Brownian motion
over the unit interval $[0, 1]$, too, when taking the expected response
at each score to be equal to the score, that is, $r(s) = s$
for every score $s$.
\end{theorem}

@@

\begin{figure}
\begin{center}
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/kuiper.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/kuiper_log.pdf}}
\end{center}
\vspace{-.125in}
\caption{Both plots graph $1 - F(x)$ versus $x$, where
$F$ is defined in~(\ref{cdf}) and is central to Corollary~\ref{corollary}.
The plot on the right uses a logarithmic scale for the vertical axis,
unlike the plot on the left. The vertical dotted line indicates
the value of $x$ corresponding to the mean of the distribution
for which $F$ is the cumulative distribution function.}
\label{kuiper_plot}
\end{figure}

@@

\begin{figure}
\begin{center}
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/gauss.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]{../codes/gauss_log.pdf}}
\end{center}
\vspace{-.125in}
\caption{Both plots graph $1 - \Phi(x)$ versus $x$,
where $\Phi$ is the cumulative distribution function
for the standard normal distribution;
$\Phi(x) = \int_{-\infty}^x \exp(-y^2/2) \, dy \, / \, \sqrt{2\pi}$.
The plot on the right uses a logarithmic scale for the vertical axis,
unlike the plot on the left.}
\label{gaussian_plot}
\end{figure}

@@

\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kuiper_ecdf_square.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kuiper_ecdf_sqrt.pdf}}
\end{center}
\vspace{-.125in}
\caption{Calibration curves (empirical cumulative distribution functions
under the null hypothesis of perfectly calibrated data) of the Kuiper P-value
for calibration for sample sizes $n =$ 100, 1,000, 10,000; in each plot,
the dashed line connects the origin $(0, 0)$ to the point $(1, 1)$
and illustrates perfect calibration, while the curve for $n =$ 10,000
is closest to perfect, $n =$ 1,000 is next closest, and $n =$ 100
is the farthest. Subfigure (a) uses scores equispaced
on the unit interval $[0, 1]$,
(b) squares each of the initially equispaced scores,
and (c) takes the square root of each of the initially equispaced scores.
The score $s$ is the predicted probability,
with the expected response $r(s) = s$ to assess calibration.
Each empirical cumulative distribution function plotted arises
from 100,000 data sets generated independently
while assuming the null hypothesis.}
\label{kuc}
\end{figure}

@@

\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kolmogorov_smirnov_ecdf_square.pdf}}
\quad
\parbox{\imsize}{\includegraphics[width=\imsize]
       {../codes/kolmogorov_smirnov_ecdf_sqrt.pdf}}
\end{center}
\vspace{-.125in}
\caption{Calibration curves (empirical cumulative distribution functions
under the null hypothesis of perfectly calibrated data)
of the Kolmogorov-Smirnov P-value for calibration
for sample sizes $n =$ 100, 1,000, 10,000; in each plot, the dashed line
connects the origin $(0, 0)$ to the point $(1, 1)$ and illustrates
perfect calibration, while the curve for $n =$ 10,000 is closest to perfect,
$n =$ 1,000 is next closest, and $n =$ 100 is the farthest.
Subfigure (a) uses scores equispaced on the unit interval $[0, 1]$,
(b) squares each of the initially equispaced scores,
and (c) takes the square root of each of the initially equispaced scores.
The score $s$ is the predicted probability,
with the expected response $r(s) = s$ to assess calibration.
Each empirical cumulative distribution function plotted arises
from 100,000 data sets generated independently
while assuming the null hypothesis.}
\label{ksc}
\end{figure}

@@

@@

\bibliography{paper}

\end{document}
